---
title: "bivariate_analysis"
author: "Moh Kyi"
format: html
editor: visual
---

## Basic commands for summary statistics

```{r}
# Load the built in dataset
data1 <- iris
```

## Summary of the data set
```{r}
summary(data1)
```

## mean of the sepal length
```{r}
mean (data1$Sepal.Length)
```
## median and sd of the sepal length
```{r}
median (data1$Sepal.Length)
sd (data1$Sepal.Length)
```
## percentile 
```{r}
quantile (data1$Sepal.Length, 0.1)
quantile (data1$Sepal.Length, c(0.1,0.9))
```
## IQR
```{r}
quantile (data1$Sepal.Length, seq(0,1,by=0.25))
```
## IQR
```{r}
IQR (data1$Sepal.Length)
```
## Some functions for data exploration
```{r}
x <- c(5,2,2,5,2,5)
```

```{r}
duplicated(x)
```

```{r}
sum(duplicated(x))
```

## Import the data file

```{r}
lbw <- rio::import("lowbwt.dta")
```

## Explore the data set

```{r}
summary(lbw) # summary statistics of the data set
```

```{r}
str(lbw) # structure of the data set
```

```{r}
dim(lbw) # dimension of the data set
```
```{r}
nrow(lbw) # number of rows
```

```{r}
ncol(lbw) # number of columns
```

## to find the duplicate 
```{r}
duplicated(lbw)
```

```{r}
sum(duplicated(lbw))
```

## Missing values
```{r}
sum(is.na(lbw))
```

```{r}
sum(is.na(lbw$id))
```

```{r}
# install.packages("mStats")
library(mStats) # You can use stata command in R
```

```{r}
codebook(lbw)
```

## mean, median,sd, min, max, range, quantile, IQR, variance (var) for birthweight (bwt)

```{r}
summary(lbw$bwt) # use base R function

```

## Using the mStats package
```{r}
summ(lbw, "bwt", "lwt", "age") # mStats function for continuous variables
```

## data visualization (exploration) of continuous variables
```{r}
## histrogram
## boxplot

hist(lbw$bwt)
boxplot(lbw$bwt)
```

## Tables
```{r}
table(lbw$smoke) # for categorical variable
```

```{r}
table(lbw$smoke, lbw$ht) # two way table
```

## factorization of the categorical variables

```{r}
lbw$low <- factor(lbw$low,
                  levels = c(0,1),
                  labels =c("BW<=2500g","BW>2500g"))
```

### Factorized the race variable (white, blck and others)
```{r}
lbw$race <- factor(lbw$race,
                   levels = c(1,2,3),
                   labels = c("white","black","others"))
```

## Table of the factorized variables
```{r}
table(lbw$race)
table(lbw$low)
```
## proportion table 
```{r}
prop.table(table(lbw$low))
prop.table(table(lbw$low))*100
round(prop.table(table(lbw$low))*100,1)
```

## Two way table 
```{r}
prop.table(table(lbw$low, lbw$race))
prop.table(table(lbw$low, lbw$race), 1) # row percentage
prop.table(table(lbw$low, lbw$race), 2) # column percentage
round(prop.table(table(lbw$low, lbw$race), 2)*100,1)
```
## Simple bar chart
```{r}
barplot(table(lbw$low))
barplot(prop.table(table(lbw$low)))
barplot(prop.table(table(lbw$low))*100)
```

```{r}
mosaicplot((~lbw$race+lbw$low), color=TRUE)
```
## using the mStats package

```{r}
mStats::tab(lbw, "low") # one way table for categorical variable
mStats::tab(lbw,"race") # one way table for categorical variable
mStats::tab(lbw,"low", by="race") # two way table for categorical variable
```

