---
title: "data_viz"
author: "Moh_kyi"
format: html
editor: visual
---

# Descriptive Statistics and Data Visualization

# Install the required packages and load theie library

```{r}
library(tidyverse)
library(dplyr)
library(ggthemes)
library(palmerpenguins)
```

# Load the data

```{r}
data(penguins)
```

# View the structure of the data

```{r}
head(penguins)
summary(penguins)
str(penguins)
glimpse(penguins)
```

## Data visualization

```{r}
# step 1
ggplot(data=penguins)
```

```{r}
# step 2
ggplot(data=penguins,
       mapping=aes(x=flipper_length_mm,
                   y=body_mass_g))
```

```{r}
# step 3
ggplot(data=penguins,
       mapping=aes(x=flipper_length_mm,
                   y=body_mass_g)) +
  geom_point()
```

```{r}
# step 4
ggplot(data=penguins,
       mapping = aes(x=flipper_length_mm,
                                 y=body_mass_g,
                                 color=species)) +
  geom_point()
```

```{r}
# step 5
ggplot(data=penguins,
       mapping = aes(x=flipper_length_mm,
                                 y=body_mass_g,
                                 color=species)) +
  geom_point() +
  geom_smooth(method="lm")
```

```{r}
# step 6
ggplot(data=penguins,
       mapping = aes(x=flipper_length_mm,
                                 y=body_mass_g
                                )) +
  geom_point(mapping=aes(color=species)) +
  geom_smooth(method="lm")
```

```{r}
# step 7
ggplot(data=penguins,
       mapping = aes(x=flipper_length_mm,
                                 y=body_mass_g
                                )) +
  geom_point(mapping=aes(color=species,shape= species)) +
  geom_smooth(method="lm")
```

```{r}
# step 8
ggplot(data=penguins,
       mapping = aes(x=flipper_length_mm,
                                 y=body_mass_g
                                )) +
  geom_point(mapping=aes(color=species,shape= species)) +
  geom_smooth(method="lm") +
  labs(
    title = "Body mass and flipper length", 
    subtitle = " Dimensions for Adelie, Chinstrap, and Gentoo Penguins",
    x = "Flipper length (mm)",
    y = "Body mass (g)",
    color = "Species",
    shape = "Species"
  )
```

```{r}
# step 9
ggplot(data=penguins,
       mapping = aes(x=flipper_length_mm,
                                 y=body_mass_g
                                )) +
  geom_point(mapping=aes(color=species,shape= species)) +
  geom_smooth(method="lm") +
  labs(
    title = "Body mass and flipper length", 
    subtitle = " Dimensions for Adelie, Chinstrap, and Gentoo Penguins",
    x = "Flipper length (mm)",
    y = "Body mass (g)",
    color = "Species",
    shape = "Species"
  ) +
  scale_color_colorblind()
```

```{r}
# More concise code using the pipe operator
penguins |>
  ggplot(aes(x=flipper_length_mm,
             y=body_mass_g)) +
  geom_point(aes(color= species, shape= species)) +
  geom_smooth(method="lm") +
  labs(
    title = "Body mass and flipper length",
    subtitle = " Dimensions for Adelie, Chinstrap, and Gentoo Penguins",
    x = "Flipper length (mm)",
    y = "Body mass (g)"
  ) +
  scale_color_colorblind()

# pipe operator %>% , is used to pass the penguins data frame to ggplot function
```

## Categorical variables
```{r}
# Step 1
ggplot(penguins, aes(x=species)) +
  geom_bar()
```

```{r}

ggplot(penguins, aes(x=species)) +
  geom_bar(aes(color=species, fill=species)) +
  labs(
    title = "Count of penguins by species",
    x = "Species",
    y = "Count"
  ) +
  scale_color_colorblind() +
  scale_fill_colorblind()
  
```

```{r}
ggplot(penguins, aes(x=species)) +
  geom_bar(aes(color=species, fill=species)) +
  labs(
    title = "Count of penguins by species",
    x = "Species",
    y = "Count"
  ) +
  theme_minimal() 
```

```{r}
#Odering the bar
ggplot(penguins, aes(x=fct_infreq(species))) +
  geom_bar(aes(color=species, fill=species)) +
  labs(
    title = "Count of penguins by species",
    x = "Species",
    y = "Count"
  ) +
  scale_color_colorblind() +
  scale_fill_colorblind()
```

## Histograms
```{r}
ggplot(penguins, aes(x= body_mass_g)) +
  geom_histogram (binwidth = 200)
```

```{r}
ggplot(penguins, aes(x= body_mass_g)) +
  geom_histogram (binwidth = 2000)
```
  
```{r}
ggplot(penguins, aes(x= body_mass_g)) +
  geom_histogram (binwidth = 20)
```

## Density plot
```{r}
# Step 1
ggplot(penguins,
       aes(x = body_mass_g)) +
  geom_density()
```

```{r}
# Step 2
ggplot(penguins,
       aes(x = body_mass_g, color= species)) +
  geom_density(linewidth = 0.75)
```

```{r}
# Step 3
ggplot(penguins,
       aes(x = body_mass_g, color= species, fill = species)) +
  geom_density(alpha= 0.1)
```

```{r}
ggplot(penguins,
       aes(x = body_mass_g, color= species, fill = species)) +
  geom_density(alpha= 0.5)
##  alpha value of 0.5 makes the density curves more opaque, allowing for better visibility of overlapping areas between species.
```

## boxplot
```{r}
ggplot(penguins, aes(x=species, y=body_mass_g)) +
  geom_boxplot()
```

```{r}
ggplot(penguins, aes(x=species, y=body_mass_g)) +
  geom_boxplot(aes(color = species, fill = species)) +
  labs(
    title = "Boxplot of body mass by species",
    x = "Species",
    y = "Body mass (g)"
  ) +
  scale_color_colorblind()
  scale_fill_colorblind()
```

## stacked bar plot for two categorical variables
```{r}
# step 1 
ggplot(penguins, aes(x= island, fill=species)) +
  geom_bar()
```

```{r}
ggplot(penguins, aes(x= island, fill=species)) +
  geom_bar(position = "fill") +
  labs(
    y= "Proportion")
```
## side by side bar plot
```{r}
ggplot(penguins, aes(x= island, fill=species)) +
  geom_bar(position = "dodge") +
  labs(
    y= "Count")
```

## faceted side by side bar plot
```{r}
ggplot(penguins, aes(x= island, fill=species)) +
  geom_bar() +
  facet_wrap(~species)
```

```{r}
ggplot(penguins, aes(x= island, fill=species)) +
  geom_bar(position ="dodge") +
  labs(
    y= "Count") +
  facet_wrap(~species)
```

## faceted proportion plot
```{r}
ggplot(penguins, aes(x= island, fill=species)) +
  geom_bar(position ="fill") +
  labs(
    y= "proportion") +
  facet_wrap(~species)
```

## faceted proprtion plot with different scale for each faceted
```{r}
ggplot(penguins, aes(x= island, fill=species)) +
  geom_bar(position ="fill") +
  labs(
    y= "proportion") +
  facet_wrap(~species, scales = "free_y", ncol =1)
```

## faceted grid for continuous variables

```{r}
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point(aes(color = species, shape = island)) +
  facet_grid(species ~ island) 
```

## Three or more variables with faceted grid
```{r}
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point(aes(color = species, shape = island)) +
  facet_grid(sex ~ island) 
```

## Faceted grid with free scales
```{r}
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point(aes(color = species, shape = island)) +
  facet_grid(sex ~ island, scales = "free" ) 
```
## faceted grid with free scale and space
```{r}
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point(aes(color = species, shape = island)) +
  facet_grid(sex ~ island, scales = "free", space = "free") 
```
## faceted grid with free scale and space and row
```{r}
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point(aes(color = species, shape = island)) +
  facet_grid(sex ~ island, scales = "free", space = "free", switch = "y") 
```

```{r}
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point(aes(color = species, shape = island)) +
  facet_grid(sex ~ island, scales = "free", space = "free", switch = "y",
  labeller = labeller(sex= c("female"= "Female Penguins", "male"= "Male Penguins")))
```

## change the theme of the plot

```{r}
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point(aes(color = species, shape = island)) +
  facet_grid(sex ~ island, scales = "free", space = "free", switch = "y",
  labeller = labeller(sex= c("female"= "Female Penguins", "male"= "Male Penguins"))) +
  theme(strip.placement = "outside")
```

```{r}
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point(aes(color = species, shape = island)) +
  facet_grid(sex ~ island, scales = "free", space = "free", switch = "y",
  labeller = labeller(sex= c("female"= "Female Penguins", "male"= "Male Penguins"))) +
  theme(strip.background = element_rect(fill = "lightblue"))
```

```{r}
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point(aes(color = species, shape = island)) +
  facet_grid(sex ~ island, scales = "free", space = "free", switch = "y",
  labeller = labeller(sex= c("female"= "Female Penguins", "male"= "Male Penguins"))) +
  theme(strip.background = element_rect(fill = "lightblue"),
        strip.text = element_text(size=12, face = "bold", color = "darkblue"))
```

```{r}
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point(aes(color = species, shape = island)) +
  facet_grid(sex ~ island, scales = "free", space = "free", switch = "y",
  labeller = labeller(sex= c("female"= "Female Penguins", "male"= "Male Penguins"))) +
  theme(strip.background = element_rect(fill = "lightblue"),
        strip.text = element_text(size=12, face = "bold", color = "darkblue"),
        strip.text.x = element_text(size = 8, face = "bold", color = "darkred"),
        strip.text.y = element_text(size = 8, face = "bold", color = "darkgreen"))
```

## Saving the plot

```{r}
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point(aes(color = species, shape = island)) +
  facet_grid(sex ~ island, scales = "free", space = "free", switch = "y",
  labeller = labeller(sex= c("female"= "Female Penguins", "male"= "Male Penguins"))) +
  theme(strip.background = element_rect(fill = "lightblue"),
        strip.text = element_text(size=12, face = "bold", color = "darkblue"),
        strip.text.x = element_text(size = 8, face = "bold", color = "darkred"),
        strip.text.y = element_text(size = 8, face = "bold", color = "darkgreen"))
ggsave(filename = "penguin_plot.png")
```
