---
title: "Import_data"
author: "Moh_Kyi"
format: html
editor: visual
---

## Import the various types of data into R

### Requuired Packages

```{r}
#install.packages("readr","readxl","data.table","haven","foreign")

library(haven)     # For reading SPSS, Stata, and SAS files
library(foreign)   # For reading older data formats like .dbf, .dta, .sav
library(readr)      # For reading CSV files
library(readxl)     # For reading Excel files
library(data.table) # For reading large data files efficiently
library(writexl)   # For writing Excel files
library(rio)        # For importing and exporting various data formats)
```

### Import the R data file (.rds, .RData)

```{r}
# Import rds file
lbw_rdata <- readRDS("lowbwt.rds")
```

## Import the delimited text file (.csv, .txt)

```{r}
# Import the text file
lbw_txt <- data.table("lowbwt.txt")
```

```{r}
# Import the csv file
lbw_csv <- read_csv("lowbwt.csv")
```

### Import the Excel file (.xls, .xlsx)

```{r}
# Import the excel file
lbw_excel <- read_excel("lowbwt.xls")
```

### Import the SPSS file (.sav)

```{r}
# Import the SPSS file
lbw_spss <- read_sav("lowbwt.sav")
```

### Import the Stata file (.dta)

```{r}
# Import the Stata file
lbw_stata <- read_dta("lowbwt.dta")
```

### Import the SAS file (.sas7bdat)

```{r}
# Import the SAS file
lbw_sas <- read_sas("lowbwt.sas7bdat")
```

### Import the Epi Info file (.epi)

```{r}
# Import the Epi Info file
lbw_epi <- read.dbf("lowbwt.rec")
```

## Export data from R to various file formats

### save RDS file

```{r} 
lowbwt  <- read_sav("lowbwt.sav")
saveRDS(lowbwt, "lowbwt_rds2.rds")
```
### Export to CSV file
```{r}
write.csv(lowbwt, "lowbwt_revised.csv", row.names=FALSE)
```
### Export to Excel file
```{r}
write_xlsx(lowbwt, "lowbwt_revised.xlsx")
```
## Exporting sav, dta and sas file
```{r}
write_sav(lowbwt, "lowbwt_revised.sav")
write_dta(lowbwt, "lowbwt_revised.dta")
write_xpt(lowbwt, "lowbwt_revised.sas7bdat")
```

## check all the available data sets in R
```{r}
data()
cars <- data("mtcars")
write.csv(cars, "mtcars_data.csv", row.names=FALSE)
```
### check all available data sets in Base R only
```{r}
data(package="datasets")
```
### Check available dat sets in the package "Epi"
```{r}
data(package="Epi")
```
 ### Import the base R data 
```{r}
data("airquality")

```
## Rio package
```{r}
# Import data using rio
lbw_rio <- import("lowbwt.csv")
# Export data using rio
export(lbw_rio, "lowbwt_rio.xlsx")
```




