---
title: "Stat&R"
Author: "Moh-Moh Kyi"
date: "2024-06-12"
output-file: "index.html"
---

# Introduction to R Programming

R is a powerful programming language and environment for statistical computing and graphics. It is widely used among statisticians and data miners for developing statistical software and data analysis. 
## Basic Syntax Here are some basic syntax elements in R:


```{r}
# Assignment
x <- 10
y <- 20
```

# Conditional Statements
``` {r}
if (x > y) {
  print("x is greater than y")
} else {
  print("x is not greater than y")
}
```

# Loops
``` {r}
for (i in 1:5) {
  print(i)
}
```

# Functions

``` {r}
add <- function(a, b) {
  return(a + b)
}

result <- add(10, 20)
print(result)## Output: 30##
```

## Data Structures

R has several built-in data structures, including vectors, lists, matrices, data frames, and factors. Here are some examples:

``` {r}
# Vectors
x <- c(1, 2, 3, 4, 5)
print(x)

```

# Lists
``` {r}
y <- list("apple", "banana", "cherry")
print(y)
```
# Matrices
``` {r}
z <- matrix(1:6, nrow = 2, ncol = 3)
print(z)
```
# Data Frames
``` {r}
data <- data.frame(
  name = c("Alice", "Bob", "Charlie"),
  age = c(25, 30, 35),
  salary = c(50000, 60000, 70000)
)
print(data)
```
# Tibble

```{r}
library(tidyverse)
tibble_example<- as.tibble(data)
```
# Factors

```{r}
gender <- factor(c("male", "female", "male", "female"))
print(gender)
```


## Conclusion

This is a brief introduction to R programming. R has many more features and capabilities for data analysis, statistical modeling, and visualization. Explore further to unlock its full potential!

#Installinng Packages
```{r}
install.packages("ggplot2")
install.packages("dplyr")
install.packages("tidyr")
install.packages("readr")
install.packages("purrr")
install.packages("tibble")
install.packages("stringr")
install.packages("forcats")
```

# Github
To use GitHub with R, you can use the `usethis` and `devtools` packages. Here are some basic steps to get started:
```{r}
install.packages("usethis")
install.packages("devtools")
library(usethis)
library(devtools)
# Create a new GitHub repository
usethis::create_github_repo("my-new-repo")
# Clone an existing GitHub repository
usethis::use_git()# Initialize a Git repository in your R project
# Commit changes to your Git repository
usethis::use_git_commit("Initial commit")
# Push changes to GitHub
usethis::use_github()_push()
```
# Additional Resources
- [R Documentation](https://www.rdocumentation.org/)
- [R for Data Science](https://r4ds.had.co.nz/)
- [CRAN R Project](https://cran.r-project.org/)
- [RStudio](https://rstudio.com/)
- [GitHub Guides](https://guides.github.com/)
- [R Markdown](https://rmarkdown.rstudio.com/)
- [R Markdown Cheatsheet](https://rstudio.com/wp-content/uploads/2015/03/rmarkdown-cheatsheet.pdf)
- [Tidyverse](https://www.tidyverse.org/)
- [R Packages](https://r-pkgs.org/)
- [R Bloggers](https://www.r-bloggers.com/)
- [Stack Overflow R Tag](https://stackoverflow.com/questions/tagged/r)
- [DataCamp R Courses](https://www.datacamp.com/courses/tech:r)
- [Coursera R Programming Course](https://www.coursera.org/learn/r-programming)
- [edX R Courses](https://www.edx.org/learn/r-programming)
- [YouTube R Tutorials](https://www.youtube.com/results?search_query=r+programming)
- [RStudio Community](https://community.rstudio.com/)
- [R Weekly Newsletter](https://rweekly.org/)
- [R-Ladies Global](https://rladies.org/)
- [R Consortium](https://www.r-consortium.org/)

# Happy Coding!
Enjoy your journey with R programming!

---

# Introduction to R Programming
R is a powerful programming language and environment for statistical computing and graphics. It is widely used among statisticians and data miners for developing statistical software and data analysis.
## Basic Syntax
Here are some basic syntax elements in R:

```{r}
# Github Installation
devtools::install_github("waffle")
```
# Loading Packages
```{r}
library(tidyverse)
library(ggplot2)
library(psych)
```

# Importing Data
```{r}
read.csv("data/2019 English R Community Survey Responses.csv")
rsurvey_csv <- read.csv("data/2019 English R Community Survey Responses.csv")
```

# Importing Excel Data
```{r}
install.packages("readxl")
library(readxl)


read.xlsx("data/2019 English R Community Survey Responses.xlsx")
rsurvey_xlsx <- read.xlsx("data/2019 English R Community Survey Responses.xlsx")
```

# Importing Stata Data
```{r}
read.dta("data/2019 English R Community Survey Responses.dta")
rsurvey_dta <- read.dta("data/2019 English R Community Survey Responses.dta")
```

# Importing SPSS Data
```{r}
install.packages("haven")
library(haven)

read.spss("data/2019 English R Community Survey Responses.sav")
rsurvey_sav <- read.spss("data/2019 English R Community Survey Responses.sav")
```
# Importing from Internet

```{r}
rsurvey_csv_from_net <- read.csv("https://github.com/rstudio/learning-r-survey/blob/master/2019/data/2019%20English%20R%20Community%20Survey%20Responses%20.csv?raw=true")
```

# Commenting
```{r}
# This is a single line comment
/* This is a multi-line comment */
```
# Section Scripts
```{r}
# Section 1 ---
# Section 2 ---
```
# Data object name
```{r}
rsurvey_csv
rsurvey_xlsx
```
# Shortcuts
```{r}
# Ctrl + Enter : Run current line or selection
# Ctrl + Shift + C : Comment/Uncomment selected lines
# Ctrl + Shift + M : Insert pipe operator (%>%)
# Ctrl + S : Save the current script
# Ctrl + Shift + N : Create a new R script
# Ctrl + 1 : Focus on Source Editor
# Ctrl + 2 : Focus on Console
# Ctrl + L : Clear the Console
# Ctrl + Shift + F10 : Restart R session
# Ctrl + Alt + I : Insert R Markdown chunk
# Ctrl + Shift + K : Knit R Markdown document
# Ctrl + Shift + T : Open a new tab in Source Editor
# Ctrl + Shift + F : Find in files
# Ctrl + Shift + H : Set working directory
# Ctrl + Alt + B : Build and Reload package
# Ctrl + Alt + P : Preview R Markdown document
# Ctrl + Alt + R : Run all code in the current script
# assign shortcut : Alt + - (to insert assignment operator <-)
# pipe shortcut : Ctrl + Shift + M (to insert %>% operator)
# help shortcut : F1 (to open help for the function under the cursor)
# plot shortcut : Ctrl + Shift + P (to open the plot pane)
# history shortcut : Ctrl + Up Arrow (to open the command history)
# tab completion : Tab (to auto-complete function and variable names)
# multi-cursor editing : Alt + Click (to place multiple cursors for simultaneous editing)
```
# Thank you for today lesson!
See you in the next lesson!
Bye!
---


