---
title: "Stat&R"
Author: "Moh-Moh Kyi"
date: "2024-06-12"
output-file: "index.html"
---

# Introduction to R Programming

R is a powerful programming language and environment for statistical computing and graphics. It is widely used among statisticians and data miners for developing statistical software and data analysis. 
## Basic Syntax Here are some basic syntax elements in R:


```{r}
# Assignment
x <- 10
y <- 20
```

# Conditional Statements
``` {r}
if (x > y) {
  print("x is greater than y")
} else {
  print("x is not greater than y")
}
```

# Loops
``` {r}
for (i in 1:5) {
  print(i)
}
```

# Functions

``` {r}
add <- function(a, b) {
  return(a + b)
}

result <- add(10, 20)
print(result)## Output: 30##
```

## Data Structures

R has several built-in data structures, including vectors, lists, matrices, data frames, and factors. Here are some examples:

``` {r}
# Vectors
x <- c(1, 2, 3, 4, 5)
print(x)
```
# Lists
``` {r}
y <- list("apple", "banana", "cherry")
print(y)
```
# Matrices
``` {r}
z <- matrix(1:6, nrow = 2, ncol = 3)
print(z)
```
# Data Frames
``` {r}
data <- data.frame(
  name = c("Alice", "Bob", "Charlie"),
  age = c(25, 30, 35),
  salary = c(50000, 60000, 70000)
)
print(data)
```
# Factors
``` {r}
gender <- factor(c("male", "female", "male", "female"))
print(gender)
```


## Conclusion

This is a brief introduction to R programming. R has many more features and capabilities for data analysis, statistical modeling, and visualization. Explore further to unlock its full potential!