---
title: "main_type_Robject"
author: "Moh Kyi"
format: docx
editor: visual
---

# 5 main types of data *objects*

## 1. vector

## 2. matrix

## 3. array

## 4. list

## 5. data frame

# " Function" is a data object too.

# But unction are mainly used to execute R commands and not use as primary.

# 5 actions:

## 1. create component of the data object

## 2. name components of data object

## 3. index( subset) component of

## 4. replace the component

## 5. operate the components of

# Vector

## A vector is a collection of the *same elements*

```{r}
x <- c(1,2,3,4,5,6,7,8,9,10)
x
class(x)
length(x)
```

# x is a numeric vector of length 10

## character

```{r}
animal <- c("dog", "cat", "ant", "monkey")
class(animal)
length(animal)
```

## logical

```{r}
bo <- c(TRUE,FALSE, T, F, F, T, F, T)
class(bo)
length(bo)
```

## Matrix

## A matrix is a 2- dimension representation of a vector

```{r}
y <- matrix(x, nrow=3); y
y <- matrix(x, nrow=2); y
y <- matrix(x, nrow=4); y
```

# Array

## Array is an n dimensional representation of a vector

```{r}
z <- array(x, dim=c(2,3,2)); z

```

```{r}
z <- array(x, dim=c(2,3,2,2)); z
```

# List

## List is a collection of columns, each contain any kind of R object

```{r}
mylist <- list(x,y,z,animal,bo)
class(mylist)
```

# Data Frame

## data frame is a tabulr data set. it is a list in tabular form

## Each row is record/ observation

## Each Column variables/ data fields

```{r}
smoker <- c("Y", "Y", "N", "N", "Y", "Y","N", "N","Y")
lu_cancer <- c("Y","N","N","Y", "Y", "N", "Y", "Y", "N")
mydata <- data.frame(smoker,lu_cancer); mydata
class(mydata)
```

```{r}
mydata2 <-  data.frame(name=c("December","Yamin","Aung","Nyan"),
                       age=c(30,16,30,28),
                       sex=c("F","F","M", "M"))
mydata2
class(mydata2)
```

```{r}
mydata3 <- c(smoker,lu_cancer)
mydata3
class(mydata3)
```

# Some more elaboration

## Numeric

```{r}
x1 <- seq(1,12,1.0)
x1
class(x1)
# 
```

```{r}
x2 <- c(1L,12L)
x2
class(x2)

```

```{r}
sex=c("F","F","M", "M")
class(sex)
```

```{r}
edu <- c("illitrate", "read and write", "primary","secondary")
class(edu)
edu
edu_factorize <- factor(edu)
class(edu_factorize)
edu_factorize
```

```{r}
status <- c("good","better","excellent","good","better","good")
class(status)
status
```

Factorized

```{r}
status1 <- ordered(status); status1
```

```{r}
status2 <- ordered(status, levels=c("good","better","excellent"));status2
```
